Generating Random Logic Programs Using Constraint Programming

### 2020-05-16 text/paper2/paper.tex
Improve introduction

%The abstract should briefly summarize the contents of the paper in
%150--250 words.
\begin{abstract}
  We present a novel approach to generating random logic programs and
  random probabilistic logic programs using constraint programming. The
  generated programs are useful in empirical testing of inference algorithms,
  random data generation, and program learning. This approach has a major
  advantage in that one can easily add additional conditions for the generated
  programs. As an example of this, we introduce a new constraint for predicate
  independence with efficient propagation and entailment algorithms, allowing
  one to generate programs that have a certain independence structure. Finally,
  we provide a combinatorial argument for the correctness of the model and
  describe how the parameters of the model affect the empirical difficulty of
  the program generation task.

  \keywords{Constraint programming \and Probabilistic logic programming \and
    Statistical relational learning.}
\end{abstract}
% TODO: update the introduction and the abstract

@@

TODO: Review \cite{DBLP:journals/tplp/WangWM15,DBLP:conf/ijcai/AmendolaRT17,DBLP:journals/ai/AmendolaRT20}

Unifying logic and probability is a long-standing challenge in artificial
intelligence \cite{DBLP:journals/cacm/Russell15}, and, in that regard,
statistical relational learning (SRL) has developed into an exciting area that
mixes machine learning and symbolic (logical and relational) structures. In
particular, probabilistic logic programs---including languages such as PRISM
\cite{DBLP:conf/ijcai/SatoK97}, ICL \cite{DBLP:journals/ai/Poole97}, and ProbLog
\cite{DBLP:conf/ijcai/RaedtKT07}---are promising frameworks for codifying
complex SRL models. With the enhanced structure, however, inference becomes more
challenging as algorithms have to correctly handle hard and soft logical
constraints. At the moment, we have no precise way of evaluating and comparing
different inference algorithms. Incidentally, if one were to survey the
literature, one often finds that an inference algorithm is only tested on 1--4
data sets
\cite{DBLP:conf/ecai/BruynoogheMKGVJR10,DBLP:journals/tplp/KimmigDRCR11,DBLP:conf/ijcai/VlasselaerBKMR15},
originating from areas such as social networks, citation patterns, and
biological data. But how confidently can we claim that an algorithm works well
if it is only tested on a few types of problems?

About thirty years ago, SAT solving technology was dealing with a similar lack
of clarity \cite{DBLP:journals/ai/SelmanML96}. This changed with the study of
generating random SAT instances against different input parameters (e.g.,
clause length and the total number of variables) to better understand the
behaviour of algorithms and their ability to solve random synthetic problems.
Unfortunately, in the context of probabilistic logic programming, most current
approaches to random instance generation are very restrictive, e.g., limited to
clauses with only two literals \cite{DBLP:conf/lpnmr/NamasivayamT09}, or clauses
of the form $a \gets \neg b$ \cite{DBLP:journals/tocl/WenWSL16}, although some
are more expressive, e.g., defining a program only by the (maximum) number of
atoms in the body and the total number of rules \cite{DBLP:conf/iclp/ZhaoL03}.

In this work, we introduce a constraint-based representation for logic programs
based on some simple parameters that describe the program's size, what
predicates and constants it uses, etc. This representation takes the form of a
constraint satisfaction problem, i.e., a set of discrete variables and
restrictions on what values they can take. Every solution to this problem (as
output by a constraint solver) directly translates into a logic program. One can
either use random value-ordering heuristics and restarts to generate random
programs or find all (sufficiently small) programs that satisfy the given
requirements. In fact, the same model can generate both probabilistic programs
in the syntax of ProbLog \cite{DBLP:conf/ijcai/RaedtKT07} and non-probabilistic
Prolog programs.

A major advantage of a constraint-based approach is the ability to add
additional constraints as needed, and to do that efficiently (compared to
generate-and-test approaches). As an example of this, we develop a custom
constraint that, given two predicates $\mathsf{P}$ and $\mathsf{Q}$, ensures
that any ground atom with predicate $\mathsf{P}$ is independent of any ground
atom with predicate $\mathsf{Q}$. In this way, we can easily regulate the
independence structure of the underlying probability distribution. We also
present a combinatorial argument for correctness, counting the number of
programs that the model produces for various parameter values and show how the
model scales when tasked with producing more complicated programs.

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can be applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is any well-formed expression that connects
atoms using conjunction $\land$, disjunction $\lor$, and negation
$\neg$\footnote{One can add a couple of extra constraints to restrict the
  generated programs to logic programs that only allow a single disjunction with
  positive and negative literals.}.
A \emph{clause} is a pair of a \emph{head} (which is an atom) and a \emph{body}
(which is a formula). A \emph{(logic) program} is a set of clauses, and a
\emph{ProbLog program} is a set of clause-probability pairs.

In the world of constraint satisfaction, we also have \emph{(constraint)
  variables}, each with a \emph{domain}, whose values are restricted using
\emph{constraints}. All constraint variables in the model are integer or set
variables, however, if an integer refers to a logical construct
(e.g., a logical variable or a constant), we will make no distinction between
the two and often use names of logical constructs to refer to the underlying
integers. We say that a constraint variable is \emph{(fully) determined} if its
domain (at the given moment in the execution) has exactly one value. We will
often use $\Box$ as a special domain value to indicate a `disabled' (i.e., fixed
and ignored) part of the model. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. Finally, we assume that all arrays start with index
zero.

@@

In this section, we present custom constraints for stratification and
independence. Stratification is a condition that both logic programs and
probabilistic logic programs need to satisfy to ensure a unique answer
to every query
\cite{DBLP:journals/jlp/BalbinPRM91,DBLP:conf/padl/MantadelisR17}, while
independence is a fundamental concept in probability theory that imposes
structure on the probability distribution and simplifies inference. In the
context of probabilistic logic programming languages, these two seemingly
disparate concepts can be defined using the same building block, i.e., a
predicate dependency graph. Constraints are usually presented with two
algorithms: one for \emph{entailment} checking, i.e., checking if the (partially
solved) constraint model satisfies the constraint or not, and one for
\emph{propagation}, i.e., removing values from domains that are incompatible
with the current state of the model. We present both algorithms for independence
while settling for a short description of entailment for stratification.

Let $\mathscr{P}$ be a probabilistic logic program with predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with $\predicates{}$ as nodes and an edge between $\mathsf{P},
\mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$ with
$\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that the
edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the head
and at least one instance of $\mathsf{P}$ at the body such that the path from
the root to the $\mathsf{P}$ node in the tree representation of the clause
passes through at least one negation node; otherwise, it is \emph{positive}. We
say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. A program
$\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$ has no negative cycles.
Thus a simple entailment algorithm for stratification can be constructed by
selecting all clauses, all predicates of which are fully determined, and looking
for negative cycles in the dependency graph constructed based on those clauses
using an algorithm such as Bellman-Ford.

@@

\section{Experimental Results}

This section presents two empirical investigations: in \cref{sec:experiment1} we
examine the scalability of our constraint model with respect to its parameters
and in \cref{sec:experiment2} we demonstrate how the model can be used to
compare inference algorithms and describe their behaviour across a wide range of
programs---something that has not been done before. The experiments were run on
a system with Intel Core i5-8250U processor, \SI{8}{\giga\byte} RAM, and Arch
Linux~5.6.11-arch1-1 operating system. The constraint model was implemented in
Java~8 with Choco~4.10.2. All inference algorithms are implemented in
ProbLog~2.1.0.39 and were run using Python~3.8.2 with PySDD~0.2.10 and
PyEDA~0.28.0. For both sets of experiments, we generate programs without
negative cycles and use a \SI{60}{\second} timeout.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups, and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into:
\begin{enumerate*}
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.}. We ran close to \num{400000} experiments, investigating
whether the model is efficient enough to generate reasonably-sized programs and
gaining insight into what parameter values make the constraint satisfaction
problem harder. For $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$,
$\maxNumNodes{}$, and $\maxNumClauses{} - |\predicates{}|$ (i.e., the number of
clauses in addition to the mandatory $|\predicates{}|$ clauses), we assign all
combinations of 1, 2, 4, 8. $\maxArity{}$ is assigned to values 1--4. For each
$|\predicates{}|$, we also iterate over all possible numbers of independent
pairs of predicates, ranging from 0 up to $\binom{|\predicates{}|}{2}$. For each
combination of the above-mentioned parameters, we pick ten random ways to assign
arities to predicates (such that $\maxArity{}$ occurs at least once) and ten
random combinations of independent pairs.

@@

\begin{figure}[t]
  \centering
  \input{line_plots.tex}%
  \caption{Mean inference time for a range of ProbLog inference algorithms as
    a function of the total number of facts in the program and the proportion of
    independent pairs of predicates. For the second plot, the number of facts is
    fixed at $10^5$.}
  \label{fig:line_plots}
\end{figure}

>>>>

%The abstract should briefly summarize the contents of the paper in
%150--250 words.
\begin{abstract}
  We present a novel approach to generating random logic programs and
  random probabilistic logic programs using constraint programming. The
  generated programs are useful in empirical testing of inference algorithms,
  random data generation, and program learning. This approach has a major
  advantage in that one can easily add additional conditions for the generated
  programs. As an example of this, we introduce a new constraint for predicate
  independence with efficient propagation and entailment algorithms, allowing
  one to generate programs that have a certain independence structure. Finally,
  we provide a combinatorial argument for the correctness of the model,
  describe how the parameters of the model affect the empirical difficulty of
  the program generation task, and use the model to compare probabilistic
  inference algorithms across a range of synthetic problems.

  \keywords{Constraint programming \and Probabilistic logic programming \and
    Statistical relational learning.}
\end{abstract}

% TODO: proofread, run a grammar checker on everything

@@

Unifying logic and probability is a long-standing challenge in artificial
intelligence \cite{DBLP:journals/cacm/Russell15}, and, in that regard,
statistical relational learning (SRL) has developed into an exciting area that
mixes machine learning and symbolic (logical and relational) structures. In
particular, probabilistic logic programs---including languages such as
\textsc{PRISM} \cite{DBLP:conf/ijcai/SatoK97}, \textsc{ICL}
\cite{DBLP:journals/ai/Poole97}, and \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07}---are promising frameworks for codifying
complex SRL models. With the enhanced structure, however, inference becomes more
challenging as algorithms have to correctly handle hard and soft logical
constraints. At the moment, we have no precise way of evaluating and comparing
different inference algorithms. Incidentally, if one were to survey the
literature, one often finds that an inference algorithm is only tested on 1--4
data sets
\cite{DBLP:conf/ecai/BruynoogheMKGVJR10,DBLP:journals/tplp/KimmigDRCR11,DBLP:conf/ijcai/VlasselaerBKMR15},
originating from areas such as social networks, citation patterns, and
biological data. But how confidently can we claim that an algorithm works well
if it is only tested on a few types of problems?

About thirty years ago, SAT solving technology was dealing with a similar lack
of clarity \cite{DBLP:journals/ai/SelmanML96}. This changed with the study of
generating random SAT instances against different input parameters (e.g.,
clause length and the total number of variables) to better understand the
behaviour of algorithms and their ability to solve random synthetic problems.
Unfortunately, when it comes to generating random logic programs, all approaches
so far focused exclusively on propositional programs
\cite{DBLP:conf/ijcai/AmendolaRT17,DBLP:journals/ai/AmendolaRT20,DBLP:journals/tplp/WangWM15,DBLP:conf/iclp/ZhaoL03},
often with severely limiting conditions such as two-literal clauses
\cite{DBLP:conf/iclp/Namasivayam09,DBLP:conf/lpnmr/NamasivayamT09} or clauses of
the form $a \gets \neg b$.

In \cref{sec:heads,sec:bodies,sec:variable_symmetry}, we introduce a
constraint-based representation for logic programs based on simple parameters
that describe the program's size, what predicates and constants it uses, etc.
This representation takes the form of a \emph{constraint satisfaction problem}
(CSP), i.e., a set of discrete variables and restrictions on what values they
can take. Every solution to this problem (as output by a constraint solver)
directly translates into a logic program. One can either output all
(sufficiently small) programs that satisfy the given conditions or use random
value-ordering heuristics and restarts to generate random programs. For sampling
from a uniform distribution, the CSP can be transformed into a belief network
\cite{DBLP:conf/aaai/DechterKBE02}. In fact, the same model can generate both
probabilistic programs in the syntax of \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07} and non-probabilistic \textsc{Prolog} programs.
This is both the first time the problem of generating random logic programs has
been approached in its full generality (i.e., including first-order clauses with
variables) and the first time inference algorithms for probabilistic logic
programs can be evaluated on more than a handful of instances.

A major advantage of a constraint-based approach is the ability to add
additional constraints as needed, and to do that efficiently (compared to
generate-and-test approaches). As an example of this, in \cref{sec:independence}
we develop a custom constraint that, given two predicates $\mathsf{P}$ and
$\mathsf{Q}$, ensures that any ground atom with predicate $\mathsf{P}$ is
independent of any ground atom with predicate $\mathsf{Q}$. In this way, we can
easily regulate the independence structure of the underlying probability
distribution. In \cref{sec:counting} we also present a combinatorial argument
for correctness that counts the number of programs that the model produces for
various parameter values. We end the paper with two experimental results in
\cref{sec:experiments}: one investigating how the constraint model scales when
tasked with producing more complex programs, and one showing how the model can
be used to evaluate and compare probabilistic inference algorithms.

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can be applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is any well-formed expression that connects
atoms using conjunction $\land$, disjunction $\lor$, and negation
$\neg$\footnote{One can add a couple of extra constraints to restrict the
  generated programs to logic programs that only allow a single disjunction with
  positive and negative literals.}.
A \emph{clause} is a pair of a \emph{head} (which is an atom) and a \emph{body}
(which is a formula). A \emph{(logic) program} is a set of clauses, and a
\emph{\textsc{ProbLog} program} is a set of clause-probability pairs.

In the world of CSPs, we also have \emph{(constraint) variables}, each with a
\emph{domain}, whose values are restricted using \emph{constraints}. All
constraint variables in the model are integer or set variables, however, if an
integer refers to a logical construct (e.g., a logical variable or a constant),
we will make no distinction between the two and often use names of logical
constructs to refer to the underlying integers. We say that a constraint
variable is \emph{(fully) determined} if its domain (at the given moment in the
execution) has exactly one value. We will often use $\Box$ as a special domain
value to indicate a `disabled' (i.e., fixed and ignored) part of the model. We
write $\arrayd{a}{b}{c}$ to mean that $\variable{a}$ is an array of variables of
length $b$ such that each element of $\variable{a}$ has domain $c$. Similarly,
we write $\arrayt{a}{b}{c}$ to denote an array $\variable{a}$ of length $b$ such
that each element of $\variable{a}$ has type $\variable{c}$. Finally, we assume
that all arrays start with index zero.

@@

In this section, we present custom constraints for stratification and
independence. Stratification is a condition necessary for probabilistic logic
programs \cite{DBLP:conf/padl/MantadelisR17} and often enforced on logic
programs \cite{DBLP:journals/tcs/Bidoit91} that helps to ensure a unique answer
to every query, while independence is a fundamental concept in probability
theory that imposes structure on the probability distribution and simplifies
inference. In the context of probabilistic logic programming languages, these
two seemingly disparate concepts can be defined using the same building block,
i.e., a predicate dependency graph. Constraints are usually presented with two
algorithms: one for \emph{entailment} checking, i.e., checking if the (partially
solved) constraint model satisfies the constraint or not, and one for
\emph{propagation}, i.e., removing values from domains that are incompatible
with the current state of the model. We present both algorithms for independence
while settling for a short description of entailment for stratification.

Let $\mathscr{P}$ be a probabilistic logic program with predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with $\predicates{}$ as nodes and an edge between $\mathsf{P},
\mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$ with
$\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that the
edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the head
and at least one instance of $\mathsf{P}$ at the body such that the path from
the root to the $\mathsf{P}$ node in the tree representation of the clause
passes through at least one negation node; otherwise, it is \emph{positive}. We
say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. A program
$\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$ has no negative
cycles\footnote{This definition is an extension of a well-known result for logic
programs \cite{DBLP:journals/jlp/BalbinPRM91} to probabilistic logic programs
with arbitrary complex clause bodies.}. Thus a simple entailment algorithm for
stratification can be constructed by selecting all clauses, all predicates of
which are fully determined, and looking for negative cycles in the dependency
graph constructed based on those clauses using an algorithm such as
Bellman-Ford.

@@

\section{Experimental Results} \label{sec:experiments}

This section presents two experiments: in \cref{sec:experiment1} we examine the
scalability of our constraint model with respect to its parameters and in
\cref{sec:experiment2} we demonstrate how the model can be used to compare
inference algorithms and describe their behaviour across a wide range of
programs---something that has not been done before. The experiments were run on
a system with Intel Core i5-8250U processor, \SI{8}{\giga\byte} RAM, and Arch
Linux~5.6.11-arch1-1 operating system. The constraint model was implemented in
Java~8 with \textsc{Choco}~4.10.2 \cite{choco}. All inference algorithms are
implemented in \textsc{ProbLog}~2.1.0.39 and were run using Python~3.8.2 with
PySDD~0.2.10 and PyEDA~0.28.0. For both sets of experiments, we generate
programs without negative cycles and use a \SI{60}{\second} timeout.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups, and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into:
\begin{enumerate*}
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.}. We ran close to \num{400000} experiments, investigating
whether the model is efficient enough to generate reasonably-sized programs and
gaining insight into what parameter values make the CSP harder. For
$|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$, and
$\maxNumClauses{} - |\predicates{}|$ (i.e., the number of clauses in addition to
the mandatory $|\predicates{}|$ clauses), we assign all combinations of 1, 2, 4,
8. $\maxArity{}$ is assigned to values 1--4. For each $|\predicates{}|$, we also
iterate over all possible numbers of independent pairs of predicates, ranging
from 0 up to $\binom{|\predicates{}|}{2}$. For each combination of the
above-mentioned parameters, we pick ten random ways to assign arities to
predicates (such that $\maxArity{}$ occurs at least once) and ten random
combinations of independent pairs.

@@

\begin{figure}[t]
  \centering
  \input{line_plots.tex}%
  \caption{Mean inference time for a range of \textsc{ProbLog} inference
    algorithms as a function of the total number of facts in the program and the
    proportion of independent pairs of predicates. For the second plot, the
    number of facts is fixed at $10^5$.}
  \label{fig:line_plots}
\end{figure}
